---
title: "Data Analysis Project: Armed Forces, Baby Names, and Optimization"
author: "Rupin R ReddyReddy"
date: "November 2025"
format: 
  pdf:
    fig-cap-location: top
    toc: false
fontsize: 11pt
geometry: margin=1in
---

## Armed Forces Data Wrangling Redux

```{r}
#| label: tbl-armed-forces
#| tbl-cap: "Distribution of Army Officers by Sex and Rank"
#| echo: false
#| message: false
#| warning: false

library(tidyverse)

# data from the DOD personnel spreadsheet - Army officers only
armed_forces_data = tibble(
  rank = c("O1", "O2", "O3", "O4", "O5", "O6", "O7", "O8", "O9"),
  Male = c(7122, 9650, 20986, 12350, 6939, 3161, 100, 80, 46),
  Female = c(2400, 3006, 6053, 3044, 1531, 452, 18, 8, 5)
)

# need to reshape this so I can make a frequency table
armed_forces_filtered = armed_forces_data |>
  pivot_longer(cols = c(Male, Female), names_to = "sex", values_to = "count") |>
  uncount(count)

# make the frequency table
freq_table = armed_forces_filtered |>
  count(sex, rank) |>
  pivot_wider(names_from = rank, values_from = n, values_fill = 0)

knitr::kable(freq_table)
```

Looking at Table 1, you can see how Army Officers are distributed across different ranks by sex. Male officers way outnumber female officers at every rank level, especially at the lower ranks like O1 and O2. If sex and rank were independent, we'd expect to see similar proportions of males to females at each rank, but that's not what's happening here. At O1, females are about 25% of the total (2,400 out of 9,522), but by O6 they're only 13% (452 out of 3,613). So the proportion of female officers actually decreases as rank goes up, which shows that sex and rank aren't independent - there's definitely a relationship between them.

## Popularity of Baby Names

```{r}
#| label: fig-baby-names
#| fig-cap: "Figure 1: Popularity of the Names Emma and Liam Over Time (1880-2015)"
#| echo: false
#| message: false
#| warning: false
#| fig-alt: "Line plot showing the number of babies named Emma and Liam from 1880 to 2015. Emma shows fluctuating popularity with peaks in the early 1900s and again from 2000-2015, reaching over 20,000 babies per year. Liam shows very low popularity until around 1990, then increases dramatically to become one of the most popular names by 2015, reaching approximately 18,000 babies per year. The two lines use different line styles (solid and dashed) to distinguish between the names."

library(babynames)

# filter for the two names I picked
baby_data = babynames |>
  filter(name %in% c("Emma", "Liam")) |>
  group_by(year, name) |>
  summarize(total = sum(n), .groups = "drop")

# make the plot - using linetype instead of color
ggplot(baby_data, aes(x = year, y = total, linetype = name)) +
  geom_line(linewidth = 1) +
  labs(x = "Year", y = "Number of Babies", linetype = "Name") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

Figure 1 shows how popular the names Emma and Liam have been over time. I picked these two because they're super popular now but have really different histories. Emma's been pretty consistently used since 1880, with a big spike in the early 1900s and then another huge spike starting around 2000. Liam, though, was basically nonexistent until the 90s and then just exploded in popularity. I used a line plot because it's the best way to show trends over time, and I went with different line styles (solid vs dashed) instead of colors so it would be easier to read for people who are colorblind. It's pretty cool how Emma maintained popularity for over a century while Liam went from nowhere to one of the top names in just like 25 years.

## Plotting a Mathematical Function (Box Problem)

```{r}
#| label: fig-box-volume
#| fig-cap: "Figure 2: Volume of Box as a Function of Corner Cut Size (36×48 inch paper)"
#| echo: false
#| message: false
#| warning: false
#| fig-alt: "Curve showing box volume as a function of corner cut size for a 36 by 48 inch piece of paper. The x-axis shows cut size from 0 to 18 inches, and the y-axis shows volume in cubic inches from 0 to about 7000. The curve starts at zero, increases smoothly to a maximum of approximately 7033 cubic inches at x equals 4.85 inches, then decreases smoothly back toward zero at x equals 18 inches. The curve is smooth and parabolic in shape with a clear single peak."

# volume function for 36 by 48 inch paper
box_volume = function(x) {
  x * (36 - 2*x) * (48 - 2*x)
}

# plot it
ggplot(data = NULL) +
  stat_function(fun = box_volume, xlim = c(0, 18), linewidth = 1) +
  labs(x = "Size of Cut (inches)", y = "Volume (cubic inches)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

Figure 2 shows the volume of the box based on how big of a square we cut from each corner. The maximum volume happens at x = 4.85 inches, where you get 7,033 cubic inches. The graph makes sense when you think about it - if you don't cut much, the box is super shallow and doesn't hold much. But if you cut too much, the base gets tiny and the volume drops again. At x = 0 there's no box, and at x = 18 the whole thing collapses because you've cut away the entire width. The curve peaks right in the middle where there's a balance between height and base size, which is exactly what you'd expect from the math.

## What You Feel You Have Learned So Far

Throughout this course, I've learned a lot about working with actual data and getting it cleaned up properly using dplyr and other tidyverse tools. Before taking this class, I honestly didn't realize how much work goes into just preparing data before you can even start analyzing it. I've also gotten way better at making visualizations with ggplot2, and now I actually think about things like making sure my plots are readable for everyone, including people who might be colorblind. Probably the most important thing I've learned is to think more carefully about statistical concepts like independence between variables - now instead of just looking at numbers, I try to understand what they actually mean and whether the patterns I'm seeing are real or just random.

\newpage

## Appendix: Code

### Armed Forces Data Wrangling

```{r}
#| label: armed-forces-code
#| eval: false
#| echo: true

library(tidyverse)

# This is the data from the DOD spreadsheet we used in class
# Just focusing on Army officers (O1 through O9)
armed_forces_data = tibble(
  rank = c("O1", "O2", "O3", "O4", "O5", "O6", "O7", "O8", "O9"),
  Male = c(7122, 9650, 20986, 12350, 6939, 3161, 100, 80, 46),
  Female = c(2400, 3006, 6053, 3044, 1531, 452, 18, 8, 5)
)

# pivot_longer changes it from wide to long format
# then uncount expands it so each person is a row
armed_forces_filtered = armed_forces_data |>
  pivot_longer(cols = c(Male, Female), names_to = "sex", values_to = "count") |>
  uncount(count)

# now make the frequency table
# count gives me the frequencies, then pivot_wider makes it look like a normal table
freq_table = armed_forces_filtered |>
  count(sex, rank) |>
  pivot_wider(names_from = rank, values_from = n, values_fill = 0)

knitr::kable(freq_table)
```

### Baby Names Visualization

```{r}
#| label: baby-names-code
#| eval: false
#| echo: true
#| fig-alt: "Line plot showing the number of babies named Emma and Liam from 1880 to 2015. Emma shows fluctuating popularity with peaks in the early 1900s and again from 2000-2015, reaching over 20,000 babies per year. Liam shows very low popularity until around 1990, then increases dramatically to become one of the most popular names by 2015, reaching approximately 18,000 babies per year. The two lines use different line styles (solid and dashed) to distinguish between the names."

library(babynames)

# filter for Emma and Liam
# group_by and summarize adds up the male and female counts for each name
baby_data = babynames |>
  filter(name %in% c("Emma", "Liam")) |>
  group_by(year, name) |>
  summarize(total = sum(n), .groups = "drop")

# make the time series plot
# using linetype = name so it's accessible for colorblind people
ggplot(baby_data, aes(x = year, y = total, linetype = name)) +
  geom_line(linewidth = 1) +
  labs(x = "Year", y = "Number of Babies", linetype = "Name") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

### Box Problem Function and Plot

```{r}
#| label: box-problem-code
#| eval: false
#| echo: true
#| fig-alt: "Curve showing box volume as a function of corner cut size for a 36 by 48 inch piece of paper. The x-axis shows cut size from 0 to 18 inches, and the y-axis shows volume in cubic inches from 0 to about 7000. The curve starts at zero, increases smoothly to a maximum of approximately 7033 cubic inches at x equals 4.85 inches, then decreases smoothly back toward zero at x equals 18 inches. The curve is smooth and parabolic in shape with a clear single peak."

# define the volume function
# x is the size of the cut, then the dimensions are 36-2x and 48-2x
box_volume = function(x) {
  x * (36 - 2*x) * (48 - 2*x)
}

# plot using stat_function
# xlim goes from 0 to 18 because you can't cut more than half of 36
ggplot(data = NULL) +
  stat_function(fun = box_volume, xlim = c(0, 18), linewidth = 1) +
  labs(x = "Size of Cut (inches)", y = "Volume (cubic inches)") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```